---
title: "220420_MicroPouch_Cleaning_metadata"
author: "Sebastian Mølvang Dall"
date: "20/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Data

```{r}
library(tidyverse)
library(janitor)
library(ggpubr)
```



## Redcap metadata

```{r}
metadata <- read_csv("../data/metadata/MicroPouch_DATA_2022-04-20_1512.csv")
```


### unselecting unuseful columns

* inclusion and exclusion criteria are ofcourse checked and unchecked for all.
* pathogenic bacteria all 0

```{r}
m_c <- metadata %>%
  select(!starts_with("inclusion"), #Inclusion criteria
         !starts_with("exclusion"), #Exclusion criteria
         !starts_with("intest"), #Pathogenic bacteria
         !excluded, #All zero
         !Complete, #All "2"
         !randomization_form_complete
         
         
         )
```


Removing any numerical column with no variation

```{r}
index <- c()


for (c in colnames(metadata)) {
  tmp <- metadata %>%
    select(c(c)) %>%
    as.data.frame()
  if (!is.numeric(tmp[,1])) {
    index = c(index, c)
  }
  
  if (is.numeric(tmp[,1])) {
    
    l <- length(tmp[!is.na(tmp[,1]), 1])
    if (l > 1) {
      if (var(tmp, na.rm = T) > 0) {
      index = c(index, c)
    }
    }
    
  }
  
}



m <- metadata %>%
  select(all_of(index)) %>%
  mutate(study_id = paste0("pt",study_id))

```

### Miniman dataset
```{r}
metadata %>%
  select(!starts_with("inclusion") & 
         
         
         
         
         #Inclusion criteria
         !starts_with("exclusion") & #Exclusion criteria
         !starts_with("intest") &#Pathogenic bacteria
         !excluded & #All zero
         !randomization_form_complete &
         !family_history:patient_demografi_complete &
         !randomization_form_complete:blodprve_complete &
         !fcesprve_complete &
         !fecal_biobank &
         !fces_sample_complete
         !affringsfrekvens_complete
         )
```



```{r}
m <- metadata %>%
  select(study_id, redcap_event_name, sex, bmi, hospital, randomization_group)
```




## Donor Batch Overview and Barcode


### Load Barcodes
```{r}
barcode_metadata <- read_csv("../data/metadata/csv/barcode_metadata.csv")
```



### Overview

A donor batch is associated with a date and a donor number
```{r}
DoBa_overview <- read_csv("../data/metadata/csv/DoBa_oversigt.csv")

DoBa_o_c <- DoBa_overview %>%
  mutate(Donor = gsub("MTKF ", "do", Donor),
         Donor = gsub(" ","_",Donor))

DoBa_o_c <- DoBa_overview %>%
  mutate(Donor = gsub("MTKF ", "", Donor),
         Donor = ifelse(nchar(Donor) > 3, "asym004", paste0("do",Donor)))


DoBa_o_l <- DoBa_o_c %>%
  pivot_longer(cols = !Donor, names_to = c("batch_donation_number"),
               values_to = "batch_name") %>%
  mutate(batch_donation_number = gsub(".donation", "", batch_donation_number),
         batch_donation_number = as.numeric(batch_donation_number)) %>%
  filter(!is.na(batch_name)) %>%
  rename(donor_number = Donor)

```




* Match donor names with `DoBa_o_c`
```{r}
DoBa <- barcode_metadata %>%
  select(donor_number, batch_donation_number, barcode_clean) %>%
  full_join(DoBa_o_l, by = c("donor_number", "batch_donation_number")) %>%
  filter(!is.na(batch_donation_number)) %>%
  mutate(batch_name = paste0("(",batch_name,")"))

#write_csv(DoBa, file = "../data/metadata/csv/DoBa_bo.csv")
```




## Donor and patient match overview

>**The last mutate corrects a date for do020 batch (17.09.19) corrected from (17.19.19)**


```{r}
PaDo_overview <- read_csv("../data/metadata/MicroPouch patient donor batch oversigt april2022.csv")

PaDo_c <- PaDo_overview %>%
  mutate(MicroPouch_NP = ifelse(grepl("(np)", Patient),T, F),
         MicroPouch_open = ifelse(grepl("(o)", Patient),T, F),
         ) %>%
  mutate(Patient = gsub(pattern = "[[:punct:]]np[[:punct:]]",replacement = "",x = Patient),
         Patient = gsub(pattern = "[[:punct:]]o[[:punct:]]",replacement = "",x = Patient),
         Patient = gsub(pattern = " ", replacement = "", x = Patient))

PaDo_l <- PaDo_c %>%
  pivot_longer(Donor_1_batch_1:Donor_4_batch_2, names_to = "donor_batch", values_to = "batch") %>%
  filter(!is.na(batch)) %>% 
  separate_rows(batch, sep = " ") %>%
  mutate(cat = ifelse(nchar(batch) == 10, "batch", "donor")) %>%
  pivot_wider(names_from = cat, values_from = batch) %>%
  mutate(donor = ifelse(donor != "asymp004", paste0("do",donor), "asym004"),
         study = ifelse(MicroPouch_NP, "NP",
                        ifelse(MicroPouch_open, "OP", "MP"))) %>%
  select(-c(MicroPouch_NP, MicroPouch_open)) %>%
  mutate(batch = ifelse(test = grepl("do020", donor), yes = ifelse(test = grepl("(17.19.19)", batch), yes = "(17.09.19)", no = batch), no = batch)) # Error in patient data overview.csv

#write_csv(PaDo_c,"../data/metadata/csv/PatientDonor_match.csv")

```


## Join information in DoBa and PaDo
This will create a dataframe for donor batch connection. 

```{r}
d_pado <- DoBa %>%
  rename(batch = batch_name, donor = donor_number) %>%
  full_join(PaDo_l, by = c("donor", "batch")) %>% 
  mutate(stage = str_sub(donor_batch, 1,7),
         stage = str_replace(stage, "Donor", "treatment")) %>% 
  mutate(treatment_batch_number = str_sub(donor_batch,9,-1))
```




# Extraction koncentration and completion of study.

I need to figure out who handed in a fecal sample. Those are the persons with sample_dates in the overview metadata. This has to be done for PaPr, PaOP and PaNP
Thereafter I need to join these tables, and connect them with the EXT_conc file. Lastly I need to make a plot showing each fecal sample and its concentration. This should preferably also be done for donors and screenings.

## PaPr overview

> I have changed value pt001, kontrol30dg, 05.11.1 -> 05.11.19

```{r}
PaPr_overview <- read_csv("../data/metadata/csv/PaPr_overview.csv")


PaPr_c <- PaPr_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>%
  #relocate(Kontrol_30dg, .after = `4_behandling`) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_1 = x1_behandling, 
         treatment_2 = x2_behandling,
         treatment_3 = x3_behandling,
         treatment_4 = x4_behandling,
         control_30dg = kontrol_30dg,
         control_1mth = kontrol_1md,
         control_3mth = kontrol_3md,
         control_6mth = kontrol_6md,
         control_12mth = kontrol_12md)

PaPr_l <- PaPr_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>%
  mutate(project = "MP") %>% 
  #filter(!is.na(patient_batch)) %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```




## PaOP overview

* load
* clean up patient column
* relocate kontrol_30dg
* make long format and change patient_batch.
```{r}
PaOP_overview <- read_csv("../data/metadata/csv/PaOP_overview.csv")


PaOP_c <- PaOP_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>%
  #relocate(Kontrol_30dg, .after = `4_behandling`) %>% 
  rename(kontrol_3md = `Kontrol 3mdr`) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_1 = x1_behandling, 
         treatment_2 = x2_behandling,
         treatment_3 = x3_behandling,
         treatment_4 = x4_behandling,
         control_30dg = kontrol_30dg,
         control_1mth = kontrol_1md,
         control_3mth = kontrol_3md)

PaOP_l <- PaOP_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>% 
  mutate(project = "OP") %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```



## PaNP

* load
* clean up patient column
* make long format and change patient_batch.

```{r}
PaNP_overview <- read_csv("../data/metadata/csv/PaNP_overview.csv")


PaNP_c <- PaNP_overview %>%
  mutate(patient = gsub("[a-zA-z]", "", patient),
         patient = gsub("[ -]","",patient),
         patient = paste0("pt",patient)) %>% 
  clean_names() %>% 
  rename(inclusion = inklusion, 
         treatment_1 = x1_behandling, 
         treatment_2 = x2_behandling,
         treatment_3 = x3_behandling,
         treatment_4 = x4_behandling,
         control_30dg = kontrol_30dg,
         control_1mth = kontrol_1md,
         control_3mth = kontrol_3md,
         control_6mth = kontrol_6md,
         control_12mth = kontrol_12md)

PaNP_l <- PaNP_c %>%
  pivot_longer(-patient, names_to = "stage", values_to = "patient_batch") %>%
  mutate(patient_batch = ifelse(!is.na(patient_batch), paste0("(",patient_batch,")"),patient_batch)) %>% 
  mutate(project = "NP") %>% 
  group_by(patient) %>% 
  mutate(fecal_donation_number = row_number())
```


## Joining the overview tables

Here I join the tables and filter out any patient_batch that was not received.

Change stage names to english

Add the fecal_batch_donation for later left_join

```{r}
stage_fct <- c("inclusion", 
         "treatment_1", 
         "treatment_2",
         "treatment_3",
         "treatment_4",
         "control_30dg",
         "control_1mth",
         "control_3mth",
         "control_6mth",
         "control_12mth",
         "drop_out")

d_overview <- bind_rows(PaPr_l, PaOP_l, PaNP_l) %>% 
  filter(!is.na(patient_batch)) %>% 
  mutate(stage = factor(stage, levels = stage_fct, labels = stage_fct))
```


## add FMT/placebo information

```{r}
group <- metadata %>%
  select(study_id, randomization_group) %>%
  filter(!is.na(randomization_group)) %>%
  mutate(group = ifelse(randomization_group==0, "FMT", "placebo"),
         patient = paste0("pt",study_id),
         project = "MP") %>% 
  select(-study_id, -randomization_group)
```

## Creating d_pat

ext_conc contains the extraction conc for each sample.

```{r}
ext_conc <- read_csv("../data/metagenomes_lab/220531_metagenome_libraries.csv")

d_ext <- full_join(d_overview, ext_conc, by = c("patient", "fecal_donation_number", "project")) %>% 
  mutate(patient = as.factor(patient),
         patient = fct_rev(patient),
         project = factor(project, levels = c("MP","OP","NP")),
         amount = ext_conc * 80,
         ext_conc = round(ext_conc, 2))

```



```{r}
d_pat_don <- d_ext %>%
  left_join(group, by = c("patient", "project")) %>%
  mutate(group = ifelse(project != "MP", "FMT", group))

```

#### Fill in donor batch information

```{r}

d_fill <- d_pado %>% 
  select(donor, barcode_clean, batch_donation_number, batch) %>% 
  distinct() %>% 
  rename(patient = donor,
         sample_barcode = barcode_clean,
         fecal_donation_number=batch_donation_number,
         donor_batch = batch) %>% 
  right_join(d_pat_don, by = c("sample_barcode", "patient","fecal_donation_number")) %>% 
  unite("batch", c(patient_batch, donor_batch), na.rm = T)

```



## Donor

```{r}
d_don <- ext_conc %>% 
  filter(is.na(project), !is.na(barcode)) %>% 
  mutate(project = ifelse(grepl("DoSc", sample_barcode), "screening", "FMT_batch"),
         fecal_donation_number = ifelse(is.na(fecal_donation_number), screeningsround -1, fecal_donation_number),
         patient = as.factor(patient)) 
```



### Extraction
```{r}
d_don %>% 
  filter(!is.na(tube_barcode)) %>% 
  mutate(ext_conc = round(ext_conc, 2),
         fecal_donation_number = as.factor(fecal_donation_number)) %>% 
  ggplot(aes(x = fecal_donation_number, y = patient, fill = ext_conc, label = ext_conc)) + 
  geom_tile(color = "white") + 
  geom_text(color = "white") + 
  #scale_color_manual(values = c("white", "red")) + 
  #scale_fill_manual(values = c("chocolate4", "grey69")) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  facet_grid(. ~project, scales = "free_x", space = "free_x") + 
  ggtitle("Main micropouch completion and concentration")
```

#### Illumina library
```{r}
gg_library_donor <- d_don %>% 
  filter(!is.na(tube_barcode)) %>% 
  mutate(pool_ng = round(pool_ng, 2),
         fecal_donation_number = as.factor(fecal_donation_number)) %>% 
  ggplot(aes(x = fecal_donation_number, y = patient, fill = pool_ng, label = pool_ng)) + 
  geom_tile(color = "white") + 
  geom_text(color = "white") + 
  scale_fill_gradient(low = "red4", high = "green4") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   hjust = 1)) +
  facet_grid(. ~project, scales = "free_x", space = "free_x")
```





## Adding treatment_1-4_donor_batch
```{r}
d_treat <- d_pado %>% 
  select(-c(donor_batch,batch, barcode_clean)) %>% 
  pivot_wider(names_from = treatment_batch_number, values_from = batch_donation_number) %>% 
  rename(patient = Patient,
         project = study)
```





```{r}
d_full <- d_fill %>% 
  rename(fecal_batch_date = batch) %>% 
  left_join(d_treat) %>% 
  rename(id = patient) %>% 
  mutate(project = if_else(str_detect(sample_barcode, "DoBa"), true = "donor_batch",
                           false = if_else(str_detect(sample_barcode, "DoSc"), true = "donor_screening", false = project))) %>% 
  mutate(project = if_else(str_detect(sample_barcode, "Asym"), true = "donor_batch", false = project))
```


```{r}
#write_csv(d_full, "../data/metadata/csv/sample_metadata.csv")
```


