---
title: "Alpha/Beta diversity"
author: "Sebastian Mølvang Dall"
date: '2022-08-05'
output: html_document
---

## libraries
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
# library(here)
library(mplibrary)

source(file = "./src/functions/metadata_functions.R")
source(file = "./src/functions/bioinformatics.R")
source(file = "./src/functions/gg_functions.R")
```

## metadata + mikrobiom data
```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan("species")
```


### Isolate metadata in metaphlan table
First step is isolation of samples in colnames(metaphlan) in metadata sheet. Thereafter, relevant columns are selected.

```{r}
metadata_in_metaphlan <- metadata %>%
  filter(LibID %in% colnames(metaphlan))

t_metaphlan <- transposeMetaphlan(metaphlan)
```

```{r projectmetadata}
project_filter <- "MP"

metadata_MP <- isolateProjectMetadata(metadata, project_filter = project_filter)
metadata_MP_with_data <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = project_filter)

remission_relapse <- metadata_MP_with_data %>%
  filter(!is.na(pdai_score)) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(inclusion - followup_30d > 2 & followup_30d <= 7, "remission", "relapse")) %>%
  select(id, group, remission)

metadata_MP_remission <- left_join(metadata_MP_with_data, remission_relapse)
```

Next chunck checks which samples did not make it into the analysis
```{r}
metadata_not_in_metaphlan <- metadata %>%
  filter(!LibID %in% colnames(metaphlan)) %>%
  filter(!is.na(id), !is.na(sample_barcode))
```



### Calculating species richness and shannon diversity

Species richness is defined as species with a relative abundance >0. Alpha diversity was calculated using Shannon Index.

```{r}
# Species richness
species_richness <- calculateSpeciesRichness(metaphlan, filter_species = 0)

# Shannon diversity
shannon_diversity <- tibble(LibID = rownames(t_metaphlan), shannon = diversity(as.matrix(t_metaphlan), index = "shannon"))

alpha_diversity_df <- full_join(species_richness, shannon_diversity)
```




#### Joining with metadata_in_metaphlan
```{r}
metadata_with_alpha <- metadata_in_metaphlan %>%
  left_join(alpha_diversity_df)

metadata_MP_remission_alpha <- left_join(metadata_MP_remission, alpha_diversity_df)
```


## GGplot Alpha
### FMT/PLACEBO
```{r}
gg_alpha_rich <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 75) +
  labs(y = "Number of Species") +
  ylim(0, 100) +
  plot_theme +
  theme(legend.position = "none")

gg_alpha_shan <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "shannon") %>%
  gg_compare_means(test = "wilcox.test", label.y = 75) +
  labs(y = "Shannon Diversity") +
  ylim(0, 4.5) +
  plot_theme +
  theme(legend.position = "none")

ggarrange(gg_alpha_rich, gg_alpha_shan, ncol = 2, labels = c("A", "B"))
# ggsave(paste0("../figures/MP_alpha_rich_shan.png"), device = "png", width = 8)
```

## Beta-Diversity
### BRAY CURTIS
Distance calculated as the median bray curtis distance from patient to set of donor samples received.

```{r}
# source("./src/functions/beta_diversity_function.R")
```


```{r}
# metadata_beta_diversity_bray <- beta_diversity(metadata_MP, t_metaphlan, transform = T, stages = c("Pre", "Post"), seed = 1)
```

```{r}
metadata_beta_bray <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", TRUE)
```

```{r}
metadata_beta_bray_FMT_and_placebo <- bind_rows(
  compareFMT2ActualDonor(metadata_MP, metadata_beta_bray),
  compareFMT2DonorNotReceived(metadata_MP, metadata_beta_bray),
  comparePlacebo2DonorNotReceived(metadata_MP, metadata_beta_bray)
)
```


```{r}
metadata_beta_summarised <- metadata_beta_bray_FMT_and_placebo %>%
  group_by(id, x_axis, actual_donor, group) %>%
  summarise(median_dissimilarity = median(dissimilarity)) %>%
  arrange(id)
```

```{r}
metadata_beta_summarised %>%
  filter(x_axis %in% c("Pre", "Post", "Donor"), group == "FMT") %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post", "Donor"))) %>%
  ggplot(aes(x = x_axis, y = 1 - median_dissimilarity, color = x_axis)) +
  geom_boxplot(outlier.shape = NA) +
  geom_line(aes(group = id), alpha = 0.6, color = "grey70", position = position_dodge(0.2)) +
  geom_point(aes(group = id), position = position_dodge(0.2), size = 2) +
  facet_grid(. ~ actual_donor) +
  plot_theme
```
```{r}
metadata_beta_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)

metadata_beta_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(filter_group = "placebo", method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)
```

```{r}
gg_beta <- function(metadata_beta_summarised, method) {
  if (method == "bray") {
    y_lab <- "Similarity to Donors (Bray-Curtis)"
  } else if (method == "sorensen") {
    y_lab <- "Similarity to Donors (Sørensen Coefficient)"
  } else {
    stopifnot(method %in% c("sorensen", "bray"), "method not valid.")
  }

  gg <- metadata_beta_summarised %>%
    ggplot(aes(x = x_axis, y = 1 - median_dissimilarity, color = x_axis)) +
    geom_boxplot(outlier.shape = NA) +
    geom_line(aes(group = id), alpha = 0.6, color = "grey70", position = position_dodge(0.2)) +
    geom_point(aes(group = id), position = position_dodge(0.2), size = 2) +
    labs(y = y_lab, x = "") +
    facet_grid(. ~ paste0(group, " - ", actual_donor)) +
    plot_theme

  return(gg)
}
```

```{r}
donor_id <- metadata_MP %>%
  filter(x_axis == "Donor") %>%
  select(LibID, id) %>%
  rename(
    donor_id = id,
    donor_comparison = LibID
  )

usedLabels <- stages_labels[stages_all %in% stages]

metadata_beta_diversity_bray_summarised <- summariseBetaDiversityOutput(metadata_beta_diversity_bray, donor_id) %>%
  mutate(x_axis = factor(x_axis, levels = stages, labels = usedLabels))

if (project_filter == "MP") {
  gg_beta_comparison_FMT <- metadata_beta_diversity_bray_summarised %>%
    filter(group == "FMT") %>%
    gg_beta("bray") +
    theme(legend.position = "none")

  gg_beta_comparison_placebo <- metadata_beta_diversity_bray_summarised %>%
    filter(group == "placebo") %>%
    gg_beta("bray") +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )

  gg_bray <- ggarrange(gg_beta_comparison_FMT, gg_beta_comparison_placebo, ncol = 2, widths = c(2, 1))
  # ggsave(filename = "../figures/beta-diversity_hellinger_bray_coeff_placebo.png", device = "png", height = 5, width = 3.5)
} else if (project_filter == "NP") {
  gg_beta_comparison_NP_bray <- metadata_beta_diversity_bray_summarised %>%
    filter(group == "FMT") %>%
    gg_beta("bray", project_filter = project_filter) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.35),
      legend.position = "right"
    )
}
```

```{r}
y.lab <- "Similarity to Donors (Bray-Curtis)"

metadata_beta_diversity_bray_summarised_remission <- metadata_beta_diversity_bray_summarised %>%
  left_join(remission_relapse)
fmt_relapse <- metadata_beta_diversity_bray_summarised_remission %>%
  filter(remission == "relapse", group == "FMT", comparison == "Actual Donors")

gg_beta_remission_fmt <- metadata_beta_diversity_bray_summarised_remission %>%
  filter(group == "FMT", comparison == "Actual Donors") %>%
  ggplot(aes(x = x_axis, y = 1 - beta_diversity, color = x_axis)) +
  geom_boxplot(data = fmt_relapse, outlier.shape = NA) +
  geom_line(aes(group = id), alpha = 0.6, color = "grey70", position = position_dodge(0.2)) +
  geom_point(aes(group = id), position = position_dodge(0.2), size = 2) +
  labs(title = "FMT group", x = "", y = y.lab) +
  ylim(c(0, 0.45)) +
  facet_grid(comparison ~ remission) +
  plot_theme

gg_beta_remission_placebo <- metadata_beta_diversity_bray_summarised_remission %>%
  filter(group == "placebo", comparison == "All Donors") %>%
  ggplot(aes(x = x_axis, y = 1 - beta_diversity, color = x_axis)) +
  geom_boxplot(outlier.shape = NA) +
  geom_line(aes(group = id), alpha = 0.6, color = "grey70", position = position_dodge(0.2)) +
  geom_point(aes(group = id), position = position_dodge(0.2), size = 2) +
  labs(title = "Placebo group", x = "", y = "") +
  ylim(c(0, 0.45)) +
  facet_grid(comparison ~ remission) +
  plot_theme

ggarrange(gg_beta_remission_fmt, gg_beta_remission_placebo)
ggsave("../figures/beta-diversity_bray_coeff_remission.png", width = 8, height = 6)
```




### Sorensen

```{r}
metadata_beta_diversity_sorensen <- beta_diversity(metadata_MP, t_metaphlan, method = "sorensen", stages = stages, seed = 1)

metadata_beta_diversity_sorensen_summarised <- summariseBetaDiversityOutput(metadata_beta_diversity_sorensen, donor_id) %>%
  mutate(x_axis = factor(x_axis, levels = stages, labels = usedLabels))


if (project_filter == "MP") {
  gg_beta_comparison_FMT <- metadata_beta_diversity_sorensen_summarised %>%
    filter(group == "FMT") %>%
    gg_beta("sorensen") +
    theme(legend.position = "none")

  gg_beta_comparison_placebo <- metadata_beta_diversity_sorensen_summarised %>%
    filter(group == "placebo") %>%
    gg_beta("sorensen") +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    )

  gg_sor <- ggarrange(gg_beta_comparison_FMT, gg_beta_comparison_placebo, ncol = 2, widths = c(2, 1))
  # ggsave(filename = "../figures/beta-diversity_hellinger_bray_coeff_placebo.png", device = "png", height = 5, width = 3.5)
} else if (project_filter == "NP") {
  gg_beta_comparison_NP_sorensen <- metadata_beta_diversity_sorensen_summarised %>%
    filter(group == "FMT") %>%
    gg_beta("sorensen", project_filter = project_filter) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.35))
}
```


```{r}

if (project_filter == "MP") {
  ggarrange(gg_sor, gg_bray, ncol = 2, labels = c("A", "B"))
  # ggsave(filename = paste0("../figures/beta-diversity_median_sor_bray_alldonor_",test,".png"), device = "png", width = 10, height = 5)
} else if (project_filter == "NP") {
  widths <- c(1, 1.2)
  ggarrange(gg_beta_comparison_NP_sorensen, gg_beta_comparison_NP_bray, ncol = 2, labels = c("A", "B"), widths = widths)
  # ggsave(paste0("../figures/",project_filter,"_beta_diversity.png"), device = "png", height = 5, width = 8)
}
```







#### BRAY without Transformation
Produces supplementary figure in MicroPouch article. Bray Curtis similarity without Hellinger transformation.

```{r}
metadata_beta_diversity_bray_notrans <- beta_diversity(metadata_MP, t_metaphlan, method = "bray", transform = F, stages = stages, seed = 1)
```

```{r}
metadata_beta_diversity_bray_summarised_no_trans <- summariseBetaDiversityOutput(metadata_beta_diversity_bray_notrans, donor_id) %>%
  mutate(x_axis = factor(x_axis, levels = stages, labels = usedLabels))


gg_beta_comparison_FMT <- metadata_beta_diversity_bray_summarised_no_trans %>%
  filter(group == "FMT") %>%
  gg_beta("bray") +
  theme(legend.position = "none")

gg_beta_comparison_placebo <- metadata_beta_diversity_bray_summarised_no_trans %>%
  filter(group == "placebo") %>%
  gg_beta("bray") +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

gg_bray <- ggarrange(gg_beta_comparison_FMT, gg_beta_comparison_placebo, ncol = 2, widths = c(2, 1))
gg_bray

# ggsave(filename = "../figures/beta-diversity_median_bray_coeff_transform_F_wilcox.test.png", device = "png", width = 6, height = 5)
```

