---
title: "Alpha/Beta diversity"
author: "Sebastian Mølvang Dall"
date: '2022-08-05'
output: html_document
---

## libraries
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
# library(here)
library(mplibrary)

source(file = "./src/functions/metadata_functions.R")
source(file = "./src/functions/bioinformatics.R")
source(file = "./src/functions/gg_functions.R")
```

## metadata + mikrobiom data
```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan("species")
```


### Isolate metadata in metaphlan table
First step is isolation of samples in colnames(metaphlan) in metadata sheet. Thereafter, relevant columns are selected.

```{r}
metadata_in_metaphlan <- metadata %>%
  filter(LibID %in% colnames(metaphlan))

t_metaphlan <- transposeMetaphlan(metaphlan)
```

```{r projectmetadata}
project_filter <- "MP"

metadata_MP <- isolateProjectMetadata(metadata, project_filter = project_filter)
metadata_MP_with_data <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = project_filter)

remission_relapse <- metadata_MP_with_data %>%
  filter(!is.na(pdai_score)) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(inclusion - followup_30d > 2 & followup_30d <= 7, "remission", "relapse")) %>%
  select(id, group, remission)

metadata_MP_remission <- left_join(metadata_MP_with_data, remission_relapse)
```

Next chunck checks which samples did not make it into the analysis
```{r}
metadata_not_in_metaphlan <- metadata %>%
  filter(!LibID %in% colnames(metaphlan)) %>%
  filter(!is.na(id), !is.na(sample_barcode))
```



### Calculating species richness and shannon diversity

Species richness is defined as species with a relative abundance >0. Alpha diversity was calculated using Shannon Index.

```{r}
# Species richness
species_richness <- calculateSpeciesRichness(metaphlan, filter_species = 0)

# Shannon diversity
shannon_diversity <- tibble(LibID = rownames(t_metaphlan), shannon = diversity(as.matrix(t_metaphlan), index = "shannon"))

alpha_diversity_df <- full_join(species_richness, shannon_diversity)



metadata_with_alpha <- metadata_in_metaphlan %>%
  left_join(alpha_diversity_df)

metadata_MP_remission_alpha <- left_join(metadata_MP_remission, alpha_diversity_df)
```


## GGplot Alpha
### FMT/PLACEBO
```{r}
gg_alpha_rich <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 75) +
  labs(y = "Number of Species") +
  ylim(0, 100)

gg_alpha_shan <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "shannon") %>%
  gg_compare_means(test = "wilcox.test", label.y = 3) +
  labs(y = "Shannon Diversity") +
  ylim(0, 4.5)

ggarrange(gg_alpha_rich, gg_alpha_shan, ncol = 2, labels = c("A", "B"))
# ggsave(paste0("../figures/MP_alpha_rich_shan.png"), device = "png", width = 8)
```

## Beta-Diversity
### BRAY CURTIS
Distance calculated as the median bray curtis distance from patient to set of donor samples received.

```{r}
# source("./src/functions/beta_diversity_function.R")
```


```{r}
# metadata_beta_diversity_bray <- beta_diversity(metadata_MP, t_metaphlan, transform = T, stages = c("Pre", "Post"), seed = 1)
```

```{r}
metadata_beta_bray <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", TRUE)
```

```{r}
metadata_beta_bray_FMT_and_placebo <- createANDcombineComparedOutputs(metadata_MP, metadata_beta_bray)
metadata_beta_bray_summarised <- metadata_beta_bray_FMT_and_placebo %>%
  summariseBetaDiversityOutput()
```

```{r}
metadata_beta_bray_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)
```


### Sorensen

```{r}
metadata_beta_sor_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "sorensen", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()


metadata_beta_sor_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "sorensen") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02)
```

#### BRAY without Transformation
Produces supplementary figure in MicroPouch article. Bray Curtis similarity without Hellinger transformation.

```{r}
metadata_beta_bray_untransformed_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()

metadata_beta_bray_untransformed_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02)
```
