---
title: "Alpha/Beta diversity"
author: "Sebastian Mølvang Dall"
date: '2022-08-05'
output: html_document
editor_options: 
  chunk_output_type: console
---

## libraries
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
# library(here)
library(mplibrary)

source(file = "./functions/metadata_functions.R")
source(file = "./functions/bioinformatics.R")
source(file = "./functions/gg_functions.R")
```

## metadata + mikrobiom data
```{r}
metadata <- load_metadata()

new_metadata <- read_delim("../data/metadata/metadata.csv", delim = ";", show_col_types = FALSE) %>% 
  filter(!is.na(sample_barcode)) %>% 
  distinct(sample_barcode, .keep_all = TRUE)



metaphlan <- readr::read_delim("../data/metaphlan3/MetaPhlAn3_Subsample_5000000_species.csv", delim = "\t", show_col_types = FALSE)

new_metaphlan <- readr::read_delim("../data/metaphlan3/MetaPhlAn_4.0.3_Combined_NonHuman_Subsampled_2500000_profile.txt", delim = "\t", show_col_types = FALSE, skip = 1) %>% 
  filter_taxonomy("species") 
  
colnames_old <- colnames(new_metaphlan)
colnames_new <- str_remove(colnames_old, "\\_Non.*")


new_metaphlan <- new_metaphlan %>% 
  rename_at(all_of(colnames_old), ~ colnames_new)


```


```{r}
filter_taxonomy <- function(metaphlan, taxonomy = "species", keep_unclassified = TRUE){
  tax <- tibble(
    abbr_taxonomy = c("k", "p", "c", "o", "f", "g", "s", "t"), 
    full_taxonomy = c("kingdom", "phylum", "class", "order", "family", "genus", "species", "strain")
  )
  
  tax_lvl <- filter(tax,full_taxonomy == taxonomy)$abbr_taxonomy
  next_tax_lvl <- filter(tax, abbr_taxonomy == tax$abbr_taxonomy[grep(tax_lvl, tax$abbr_taxonomy) + 1])$abbr_taxonomy
  
  
  if(keep_unclassified){
    df <- metaphlan %>% 
      filter(
        clade_name == "UNCLASSIFIED" | str_detect(clade_name, paste0(tax_lvl,"_{2}")), 
        str_detect(clade_name, paste0(next_tax_lvl,"_{2}")) == FALSE
      )
  } else {
    df <- metaphlan %>% 
      filter(
        str_detect(clade_name, paste0(tax_lvl,"_{2}")), 
        str_detect(clade_name, paste0(next_tax_lvl,"_{2}")) == FALSE
      )
  }
  
  if(taxonomy == "strain"){
    df <- metaphlan %>% 
      filter(
        clade_name == "UNCLASSIFIED" | str_detect(clade_name, paste0(tax_lvl,"_{2}"))
      )
  }
  
  
  return(
    df %>% 
      mutate(
        clade_name = str_remove(clade_name, paste0(".*",tax_lvl,"__"))
      )
  )
  
  
}


```


### Isolate metadata in metaphlan table
First step is isolation of samples in colnames(metaphlan) in metadata sheet. Thereafter, relevant columns are selected.

```{r}
metadata_in_metaphlan <- metadata %>%
  filter(LibID %in% colnames(metaphlan))

new_metadata_in_metaphlan <- new_metadata %>% 
  filter(sample_barcode %in% colnames(new_metaphlan))

t_metaphlan <- transposeMetaphlan(metaphlan)
t_new_metaphlan <- new_metaphlan %>%
        pivot_longer(-clade_name, names_to = "sample_barcode") %>%
        pivot_wider(names_from = clade_name, values_from = value) %>%
        arrange(sample_barcode) %>%
        column_to_rownames(var = "sample_barcode")

```

```{r projectmetadata}
project_filter <- "MP"

metadata_MP <- isolateProjectMetadata(metadata, project_filter = project_filter)
metadata_MP_with_data <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = project_filter)

new_metadata_MP <- isolateProjectMetadata(new_metadata)
new_metadata_MP_with_data <- isolateProjectMetadata(new_metadata_in_metaphlan, project_filter = project_filter)



remission_relapse <- metadata_MP_with_data %>%
  filter(!is.na(pdai_score)) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(inclusion - followup_30d > 2 & followup_30d <= 7, "remission", "relapse")) %>%
  select(id, group, remission)

metadata_MP_remission <- left_join(metadata_MP_with_data, remission_relapse)



new_remission_relapse <- new_metadata_MP_with_data %>% 
  filter(!is.na(pdai_score), x_axis %in% c("Pre", "Post")) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(inclusion - followup_30d > 2 & followup_30d <= 7, "remission", "relapse")) %>%
  select(id, group, remission)

new_metadata_MP_remission <- left_join(new_metadata_MP_with_data, new_remission_relapse)
```

Next chunck checks which samples did not make it into the analysis
```{r}
metadata_not_in_metaphlan <- metadata %>%
  filter(!LibID %in% colnames(metaphlan)) %>%
  filter(!is.na(id), !is.na(sample_barcode))

new_metadata_not_in_metaphlan <- new_metadata %>% 
  filter(!is.na(sample_barcode)) %>% 
  distinct(sample_barcode, .keep_all =TRUE) %>% 
  filter(!str_detect(sample_barcode, "LIB"), !sample_barcode %in% colnames(new_metaphlan))

```



### Calculating species richness and shannon diversity

Species richness is defined as species with a relative abundance >0. Alpha diversity was calculated using Shannon Index.

```{r}
# Species richness
species_richness <- calculateSpeciesRichness(metaphlan, filter_species = 0)

# Shannon diversity
shannon_diversity <- tibble(LibID = rownames(t_metaphlan), shannon = diversity(as.matrix(t_metaphlan), index = "shannon"))
alpha_diversity_df <- full_join(species_richness, shannon_diversity)

metadata_with_alpha <- metadata_in_metaphlan %>%
  left_join(alpha_diversity_df)

metadata_MP_remission_alpha <- left_join(metadata_MP_remission, alpha_diversity_df)
```


```{r}
new_species_richness <- new_metaphlan %>% 
        pivot_longer(-clade_name, names_to = "sample_barcode") %>%
        group_by(sample_barcode) %>%
        summarise(richness = sum(value > 0))


new_shannon_diversity <- tibble(sample_barcode = rownames(t_new_metaphlan), shannon = diversity(as.matrix(t_new_metaphlan), index = "shannon"))
new_alpha_diversity_df <- full_join(new_species_richness, new_shannon_diversity)

new_metadata_with_alpha <- new_metadata_MP_remission %>%
  left_join(new_alpha_diversity_df)
```

```{r}
new_metadata %>% 
  filter(stage %in% c("inclusion", "followup_30d"), project=="MP") %>% 
  group_by(stage, group) %>% 
  summarise(n())
```


## GGplot Alpha
### FMT/PLACEBO
```{r}
gg_alpha_rich <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 75) +
  labs(y = "Number of Species") +
  ylim(0, 100)


gg_alpha(new_metadata_with_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 150) +
  labs(y = "Number of Species")

gg_alpha_shan <- gg_alpha(metadata_MP_remission_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "shannon") %>%
  gg_compare_means(test = "wilcox.test", label.y = 3) +
  labs(y = "Shannon Diversity") +
  ylim(0, 4.5)

ggarrange(gg_alpha_rich, gg_alpha_shan, ncol = 2, labels = c("A", "B"))
# ggsave(paste0("../figures/MP_alpha_rich_shan.png"), device = "png", width = 8)
```


```{r}
gg_alpha(new_metadata_with_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 150) +
  labs(y = "Number of Species")

gg_alpha(new_metadata_with_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "shannon") %>%
  gg_compare_means(test = "wilcox.test", label.y = 3) +
  labs(y = "Shannon Diversity")

```


## Beta-Diversity
### BRAY CURTIS
Distance calculated as the median bray curtis distance from patient to set of donor samples received.

```{r}
# source("./src/functions/beta_diversity_function.R")
```


```{r}
# metadata_beta_diversity_bray <- beta_diversity(metadata_MP, t_metaphlan, transform = T, stages = c("Pre", "Post"), seed = 1)
```

```{r}
metadata_beta_bray <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", TRUE)

new_metadata_beta_bray <- createMetadataWBetadiversity(new_metadata_MP, t_new_metaphlan, "bray", TRUE)

```

```{r}
metadata_beta_bray_FMT_and_placebo <- createANDcombineComparedOutputs(metadata_MP, metadata_beta_bray)
metadata_beta_bray_summarised <- metadata_beta_bray_FMT_and_placebo %>%
  summariseBetaDiversityOutput()
```

```{r}
metadata_beta_bray_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)
```


### Sorensen

```{r}
metadata_beta_sor_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "sorensen", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()


metadata_beta_sor_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "sorensen") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02)
```

#### BRAY without Transformation
Produces supplementary figure in MicroPouch article. Bray Curtis similarity without Hellinger transformation.

```{r}
metadata_beta_bray_untransformed_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()

metadata_beta_bray_untransformed_summarised %>%
  filter(x_axis %in% c("Pre", "Post")) %>%
  mutate(x_axis = factor(x_axis, levels = c("Pre", "Post"))) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02)
```
