---
title: "Alpha/Beta diversity"
author: "Sebastian Mølvang Dall"
date: '2022-08-05'
output: html_document
editor_options: 
  chunk_output_type: console
---

## libraries
```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
# library(here)
library(mplibrary)

source(file = "./functions/metadata_functions.R")
source(file = "./functions/bioinformatics.R")
source(file = "./functions/gg_functions.R")
```

## metadata + mikrobiom data
```{r}
metadata <- load_metadata()

metaphlan <- read_delim(
    "../data/metaphlan3/MetaPhlAn_4.0.3_Combined_NonHuman_Subsampled_2500000_profile.txt", 
    delim = "\t", 
    show_col_types = FALSE, 
    skip = 1
  ) %>% 
    filter_taxonomy("species", keep_unclassified = FALSE) %>% 
    remove_NonHuman_from_colnames()


```



### Isolate metadata in metaphlan table
First step is isolation of samples in colnames(metaphlan) in metadata sheet. Thereafter, relevant columns are selected.

```{r}
metadata_in_metaphlan <- metadata %>% 
  filter(sample_barcode %in% colnames(metaphlan))

t_metaphlan <- metaphlan %>%
        transposeMetaphlan()

```

```{r projectmetadata}
project_filter <- "MP"

metadata_MP <- isolateProjectMetadata(metadata, project_filter = project_filter)
metadata_MP_with_data <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = project_filter)


remission_relapse <- metadata_MP_with_data %>% 
  filter(!is.na(pdai_score), x_axis %in% c("Pre", "Post")) %>%
  group_by(id) %>%
  filter(n_distinct(stage) == 2) %>%
  select(id, stage, group, pdai_score) %>%
  pivot_wider(names_from = stage, values_from = pdai_score) %>%
  mutate(remission = if_else(followup_30d < 7, "remission", "relapse")) %>%
  select(id, group, remission)

metadata_MP_remission <- left_join(metadata_MP_with_data, remission_relapse)
```

Next chunck checks which samples did not make it into the analysis
```{r}
metadata_not_in_metaphlan <- metadata %>% 
  filter(!is.na(sample_barcode)) %>% 
  #distinct(sample_barcode, .keep_all =TRUE) %>% 
  filter(!str_detect(sample_barcode, "LIB"), !sample_barcode %in% colnames(metaphlan)) %>% 
  select(id, sample_barcode, project, stage, group)

```
### Calculating species richness and shannon diversity

Species richness is defined as species with a relative abundance >0. Alpha diversity was calculated using Shannon Index.

```{r}
species_richness <- calculateSpeciesRichness(metaphlan, filter_species = 0)
shannon_diversity <- tibble(sample_barcode = rownames(t_metaphlan), shannon = diversity(as.matrix(t_metaphlan), index = "shannon"))
alpha_diversity_df <- full_join(species_richness, shannon_diversity)

metadata_with_alpha <- metadata_MP_remission %>%
  left_join(alpha_diversity_df)# %>% 
  # mutate(
  #   x_axis = factor(x_axis, levels = stages_pre_post, labels = c("Before", "After", "Donor")),
  #   group = factor(group, levels = c("FMT", "placebo"), labels = c("FMT","Placebo"))
  # )
```

```{r}
metadata %>% 
  filter(stage %in% c("inclusion", "followup_30d"), project=="MP") %>% 
  group_by(stage, group) %>% 
  summarise(n())
```


## GGplot Alpha
### FMT/PLACEBO
```{r}
facet_labels <- c("FMT", "Placebo")
names(facet_labels) <- c("FMT", "placebo")

gg_alpha_rich <- gg_alpha(metadata_with_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "richness") %>%
  gg_compare_means(test = "wilcox.test", label.y = 100) +
  labs(y = "Number of Species") +
  scale_x_discrete(labels = c("Before", "After", "Donor"))


gg_alpha_shan <- gg_alpha(metadata_with_alpha, x = c("Pre", "Post", "Donor"), alpha_metric = "shannon") %>%
  gg_compare_means(test = "wilcox.test", label.y = 3.5) +
  labs(y = "Shannon Diversity") +
  scale_x_discrete(labels = c("Before", "After", "Donor"))

ggarrange(gg_alpha_rich, gg_alpha_shan, ncol = 2, labels = c("A", "B"))
ggsave(paste0("../figures/metaphlan4/alpha_beta/alpha_shannon_diversity.png"), device = "png", width = 8)
```

```{r}
gg_followup_data <- gg_alpha(metadata_with_alpha, x = c("Pre", "Post", "followup_1m"), alpha_metric = "richness")$data 


gg_followup_data_mean <- gg_followup_data %>%
  group_by(x_axis, group) %>% 
  summarise(alpha_metric = mean(alpha_metric))

gg_followup_data_n <- gg_followup_data %>% 
  group_by(x_axis, group) %>% 
  summarise(
    n = n()
  ) %>% 
  mutate(
    label = paste0("n: ", n),
    alpha_metric = 100
  )

gg_followup_richness <- gg_followup_data %>%
  ggplot(aes(x = x_axis, y = alpha_metric)) +
  geom_boxplot(outlier.shape = NA) +
  geom_line(data = filter(gg_followup_data, x_axis != "Donor"), aes(group = id), alpha = 0.6, color = "grey60", position = position_dodge(0.2), linetype = "dashed") +
  geom_point(aes(group = id), position = position_dodge(0.2), size = 2, color = "grey80") +
  geom_label(data = gg_followup_data_n, aes(label = label)) + 
  plot_theme +
  facet_grid(. ~ group) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("Before", "After", "1Mth", "3Mth")) +
  labs(y = "Number of Species")


gg_followup_data <- gg_alpha(metadata_with_alpha, x = c("Pre", "Post", "followup_1m"), alpha_metric = "shannon")$data 


gg_followup_data_mean <- gg_followup_data %>%
  group_by(x_axis, group) %>% 
  summarise(alpha_metric = mean(alpha_metric))

gg_followup_data_n <- gg_followup_data %>% 
  group_by(x_axis, group) %>% 
  summarise(
    n = n()
  ) %>% 
  mutate(
    label = paste0("n: ", n),
    alpha_metric = 4
  )

gg_followup_shannon <- gg_followup_data %>%
  ggplot(aes(x = x_axis, y = alpha_metric)) +
  geom_boxplot(outlier.shape = NA) +
  geom_line(data = filter(gg_followup_data, x_axis != "Donor"), aes(group = id), alpha = 0.6, color = "grey60", position = position_dodge(0.2), linetype = "dashed") +
  geom_point(aes(group = id), position = position_dodge(0.2), size = 2, color = "grey80") +
  geom_label(data = gg_followup_data_n, aes(label = label)) + 
  plot_theme +
  facet_grid(. ~ group) +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("Before", "After", "1Mth", "3Mth")) +
  labs(y = "Shannon Diversity")

ggarrange(gg_followup_richness, gg_followup_shannon, ncol = 2, labels = c("A", "B"))
ggsave(paste0("../figures/metaphlan4/alpha_beta/alpha_shannon_diversity_followup.png"), device = "png", width = 12)
```




### Clinical improvement and alpha-diversity
```{r}
clinical_effect_and_alpha <- metadata_with_alpha %>% 
  filter(stage %in% c("inclusion", "followup_30d"), group == "FMT") %>% 
  select(id, stage, richness, shannon, pdai_score, remission) %>% 
  pivot_wider(names_from=stage, values_from=c(richness, shannon, pdai_score)) %>% 
  mutate(
    pdai_diff = (pdai_score_followup_30d - pdai_score_inclusion),
    richness_diff = richness_followup_30d - richness_inclusion,
    shannon_diff = shannon_followup_30d - shannon_inclusion
  ) %>% 
  select(id, pdai_diff:shannon_diff, remission) %>% 
  pivot_longer(cols = c(richness_diff, shannon_diff), names_to = "alpha", values_to="value") %>% 
  filter(!is.na(pdai_diff))

lm_fit_rich <- lm(value ~ pdai_diff, data = filter(clinical_effect_and_alpha, alpha == "richness_diff"))
lm_fit_shan <- lm(value ~ pdai_diff, data = filter(clinical_effect_and_alpha, alpha == "shannon_diff"))
summary(lm_fit_rich)
summary(lm_fit_shan)
r_squared <- summary(lm_fit_rich)$r.squared


clinical_effect_and_alpha %>% 
  mutate(
    alpha = factor(alpha, levels = c("richness_diff", "shannon_diff"), labels=c("Richness (After - Before)", "Shannon (After - Before)"))
  ) %>% 
  ggplot(aes(x = pdai_diff, y = value)) +
    geom_smooth(method="lm", color="black") +
    geom_point(size = 3, aes(color = remission)) +
    facet_grid(alpha~., scales = "free_y") +
    scale_x_continuous(breaks = function(b) seq(floor(min(b)), ceiling(max(b)), by=1)) + 
    plot_theme +
    theme(
      axis.title.x = element_text(),
      legend.position = "right"
    ) +
    scale_x_reverse() + 
    labs(x = "Change in PDAI score (Before - After)", y = "Alpha diversity", title = "Comparison of clinical effect and alpha diversity in the FMT group")

ggsave("../figures/metaphlan4/alpha_beta/alpha_v_clinical_effect.png", device= "png", width = 8)
```


## Pdai
```{r}
metadata_pdai <- metadata_MP %>% 
  select(id,project, group, stage, contains("pdai")) %>% 
  filter(project == "MP") %>% 
  select(id:pdai_h2) %>% 
  mutate(
    c = pdai_stool_frequency + pdai_rectal_bleeding + pdai_bowel_abdominal + pdai_fever,
    e = pdai_edema + pdai_granularity + pdai_friability + pdai_vp_loss,
    h = pdai_h1 + pdai_h2
  ) %>% 
  select(id:stage, c,e,h) %>% 
  filter(stage %in% c("inclusion", "followup_30d")) %>% 
  pivot_longer(cols = c:h) %>% 
  mutate(value = if_else(is.na(value), 0, value))


metadata_pdai %>% 
ggplot(aes(x=factor(stage, levels=c("inclusion", "followup_30d")), y = value, color = stage)) + 
  geom_point(size = 4,alpha = 0.8, position=position_jitter(width=0.1, height=0.01)) + 
  facet_grid(group~name)
```


## Beta-Diversity
### BRAY CURTIS
Distance calculated as the median bray curtis distance from patient to set of donor samples received.

```{r}
metadata_beta_bray <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", TRUE)

metadata_beta_bray_FMT_and_placebo <- createANDcombineComparedOutputs(metadata_MP, metadata_beta_bray)
metadata_beta_bray_summarised <- metadata_beta_bray_FMT_and_placebo %>%
  summariseBetaDiversityOutput()

```

```{r}
gg_bray <- metadata_beta_bray_summarised %>%
  filter(x_axis %in% c("Pre", "Post"), actual_donor != "Donors not received") %>%
  mutate(
    x_axis = factor(x_axis, levels = c("Pre", "Post")),
    actual_donor = if_else(actual_donor == "Actual Donors", "Donors received", actual_donor)
  ) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)+
  labs(title = "Bray Similarity") +
  scale_x_discrete(labels = c("Before", "After"))
```




### Sorensen

```{r}
metadata_beta_sor_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "sorensen", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()


gg_sor = metadata_beta_sor_summarised %>%
  filter(x_axis %in% c("Pre", "Post"), actual_donor != "Donors not received") %>%
  mutate(
    x_axis = factor(x_axis, levels = c("Pre", "Post")),
    actual_donor = if_else(actual_donor == "Actual Donors", "Donors received", actual_donor)
  ) %>% 
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "sorensen") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02) +
  labs(title = "Sørensen Similarity") +
  scale_x_discrete(labels = c("Before", "After"))
```


```{r}
ggarrange(gg_sor, gg_bray, ncol = 2, labels = c("A", "B"))
ggsave(paste0("../figures/metaphlan4/alpha_beta/bray_sor_similarity_transform-true.png"), device = "png", width = 13)
```



```{r}
gg_bray_followup <- metadata_beta_bray_summarised %>%
  filter(x_axis %in% c("Pre", "Post", "followup_1m"), actual_donor != "Donors not received") %>%
  mutate(
    x_axis = factor(x_axis, levels = c("Pre", "Post", "followup_1m")),
    actual_donor = if_else(actual_donor == "Actual Donors", "Donors received", actual_donor)
  ) %>%
  group_by(id) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, label.y = 0.45, tip.length = 0.02)+
  labs(title = "Bray Similarity") +
  scale_x_discrete(labels = c("Before", "After", "1Mth"))

gg_sor_followup = metadata_beta_sor_summarised %>%
  filter(x_axis %in% c("Pre", "Post", "followup_1m"), actual_donor != "Donors not received") %>%
  mutate(
    x_axis = factor(x_axis, levels = c("Pre", "Post", "followup_1m")),
    actual_donor = if_else(actual_donor == "Actual Donors", "Donors received", actual_donor)
  ) %>% 
  group_by(id) %>%
  gg_beta(method = "sorensen") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02) +
  labs(title = "Sørensen Similarity") +
  scale_x_discrete(labels = c("Before", "After", "1Mth"))
```

```{r}
ggarrange(gg_sor_followup, gg_bray_followup, ncol = 2, labels = c("A", "B"))
ggsave(paste0("../figures/metaphlan4/alpha_beta/bray_sor_similarity_followup_transform-true.png"), device = "png", width = 13)
```





#### BRAY without Transformation
Produces supplementary figure in MicroPouch article. Bray Curtis similarity without Hellinger transformation.

```{r}
metadata_beta_bray_untransformed_summarised <- createMetadataWBetadiversity(metadata_MP, t_metaphlan, "bray", FALSE) %>%
  createANDcombineComparedOutputs(projectmetadata = metadata_MP) %>%
  summariseBetaDiversityOutput()

metadata_beta_bray_untransformed_summarised %>%
  filter(x_axis %in% c("Pre", "Post"), actual_donor != "Donors not received") %>%
  mutate(
    x_axis = factor(x_axis, levels = c("Pre", "Post")),
    actual_donor = if_else(actual_donor == "Actual Donors", "Donors received", actual_donor)
  ) %>% 
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  gg_beta(method = "bray") +
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pre", "Post")), paired = T, tip.length = 0.02) +
  scale_x_discrete(labels = c("Before", "After"))

ggsave(paste0("../figures/metaphlan4/alpha_beta/bray_similarity_transform_false.png"), device = "png", width = 7)
```
