---
title: "heatmap"
author: "Sebastian Mølvang Dall"
date: '2022-08-10'
output: html_document
---

## Loading libraries

```{r}
library(tidyverse)
library(ggpubr)
source("functions/metadata_functions.R")
```

## Loading Data

```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan(taxonomic_level = "genus")
```

```{r}
metadata_all <- prepare_metadata(metadata)

metadata_in_metaphlan <- prepare_metadata(metadata) %>% 
  filter(LibID %in% colnames(metaphlan))


metaphlan_long <- metaphlan %>% 
  select(-NCBI_tax_id) %>% 
  pivot_longer(-clade_name, names_to = "LibID") %>% 
  arrange(LibID)
```

### Filtering for relevant patients and donors

Just as in alpha_beta_diversity.rmd, the donors are filtered such that only donors used for the available patients are used.

```{r}
stages = stages_pre_post


metadata_relevant <- IsolateProjectAndDonor(metadata_in_metaphlan, project_filter = "MP") %>% 
  mutate(sample_group = paste0(group, "_", x_axis),
         sample_group = if_else(str_detect(sample_group, "FMT_Donor"), str_remove(sample_group, "FMT_"), sample_group)) %>% 
  filter(x_axis %in% stages) %>% 
  group_by(id) %>% 
  filter(n_distinct(x_axis) == 2 | x_axis == "Donor")

```

##### NP only
```{r}
stages = stages_all

metadata_relevant <- IsolateProjectAndDonor(metadata_in_metaphlan, project_filter = "NP") %>% 
  mutate(sample_group = paste0(group, "_", x_axis),
         sample_group = if_else(str_detect(sample_group, "FMT_Donor"), str_remove(sample_group, "FMT_"), sample_group)) %>% 
  filter(x_axis %in% stages) 
```


#### Adding taxonomy to metadata table

```{r}
metadata_relevant_with_tax <- left_join(metadata_relevant, metaphlan_long)
```

##### Function for visualizing patient with donor pairings.

> Filtering for most abundant taxa could be made better. Right now it takes most abundant across patient and donors.

```{r}
source("functions/bioinformatics.R")
```

```{r}
print(unique(filter(metadata_all, project == "MP" & group == "FMT")$id))
print(unique(filter(metadata_in_metaphlan, project == "MP" & group == "FMT")$id))

PatientDonorHeatmap(metadata_with_tax = metadata_relevant_with_tax, patient = "pt001", project_filter = "MP")


PatientDonorHeatmap(metadata_relevant_with_tax, patient = "pt027", project_filter = "MP")


PatientDonorHeatmap(metadata_relevant_with_tax, patient = "pt028", project_filter = "MP")



```


## heatmap visualization

### Donor, Pre, heatmap

#### Top genera

```{r}
metadata_weigth <- metadata_relevant_with_tax %>% 
  mutate(sample_group = paste0(x_axis, " ", group),
         sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group)) %>% 
  group_by(donor_batch, sample_group) %>% 
  mutate(hellinger = sqrt(value/sum(value)),
         sum = sum(value)) %>% 
  ungroup()


top_genera_weighted <- metadata_weigth %>%
  filter(x_axis %in% c("Pre", "Donor")) %>% 
  group_by(id, sample_group, clade_name) %>% 
  summarise(hellinger = mean(hellinger)) %>% 
  group_by(sample_group) %>% 
  mutate(weight = 1/n()) %>% 
  group_by(clade_name) %>% 
  summarise(mean_hellinger = weighted.mean(hellinger, weight)) %>% 
  arrange(desc(mean_hellinger)) %>% 
  head(35)

top_genera_weighted_donor <- metadata_weigth %>%
  filter(clade_name %in% top_genera_weighted$clade_name, x_axis == "Donor") %>% 
  group_by(id, clade_name) %>% 
  summarise(hellinger = mean(hellinger)) %>% 
  group_by(clade_name) %>% 
  summarise(mean_hellinger = mean(hellinger)) %>% 
  arrange(desc(mean_hellinger))

```


```{r}
l %>% 
  group_by(group) %>% 
  mutate(n = 1/n(),
         mean = mean(value)) %>% 
  ungroup() %>% 
  summarise(mean = weighted.mean(value, n))
```


#### hclust

```{r}
t_metaphlan <- metaphlan %>% 
  select(-NCBI_tax_id) %>% 
  pivot_longer(-clade_name, names_to = "LibID") %>% 
  pivot_wider(names_from = clade_name, values_from = value) %>% 
  filter(LibID %in% metadata_weigth$LibID) %>% 
  arrange(LibID) %>% 
  column_to_rownames(var = "LibID") %>% 
  as.data.frame()



t_metaphlan_id <- metadata_weigth %>% 
  filter(str_detect(sample_group, "Pre") | sample_group == "Donor") %>% 
  arrange(LibID) %>% 
  select(id, LibID, group, sample_group, x_axis) %>% 
  distinct(LibID, .keep_all = T) %>% 
  left_join(
    rownames_to_column(t_metaphlan,"LibID")
  )
  
t_metaphlan_long_id_donor_avg <- t_metaphlan_id %>% 
  pivot_longer(!id:x_axis, names_to = "clade_name", values_to = "relative_abundance") %>% 
  group_by(id, x_axis, clade_name) %>% 
  summarise(relative_abundance = mean(relative_abundance))

t_metaphlan_wide_id_donor_avg <- t_metaphlan_long_id_donor_avg %>% 
  ungroup() %>%
  select(!x_axis) %>%
  pivot_wider(names_from = clade_name, values_from = relative_abundance) %>% 
  column_to_rownames("id")

bray_curtis_dist <- vegan::vegdist(vegan::decostand(t_metaphlan_wide_id_donor_avg, method = "hellinger"))

hclust_ward <- hclust(bray_curtis_dist, method = "ward.D2")

ward_dendogram <- as.dendrogram(hclust_ward)
ward_order <- order.dendrogram(ward_dendogram)

plot(ward_dendogram)


id_cluster_order <- t_metaphlan_long_id_donor_avg %>% 
  left_join(metadata_weigth) %>% 
  mutate(
    id = factor(
      id,
      levels = hclust_ward$labels[ward_order]
    )
  ) %>% 
  group_by(id, clade_name, sample_group) %>% 
  distinct(id, .keep_all = T)
  
```


### Donor / Pre heatmap

Pre and donor heatmap clustered with hclust ward.D2 method. donor batches have been grouped by average relative abundance.

```{r}

gg_heat_inc <- id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         x_axis = factor(x_axis, levels = c("Pre", "Donor")),
         sample_group = factor(sample_group, levels = c("Pre FMT", "Pre placebo", "Donor")),
         relative_abundance = if_else(relative_abundance < 10, round(relative_abundance, 1), floor(relative_abundance))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = relative_abundance)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          legend.position = "none")
gg_heat_inc

ggsave("../figures/heatmap_pre_donor_hellinger.png", device = "png", width = 12, height = 10)
```




### Post heatmap


```{r}

metadata_tax_post <- metadata_weigth %>% 
  filter(x_axis == "Post") %>% 
  arrange(LibID)



t_metaphlan_post_id <- metadata_tax_post %>%  
  select(id, LibID, group, sample_group,x_axis) %>% 
  distinct(LibID, .keep_all = T) %>% 
  left_join(
    rownames_to_column(t_metaphlan,"LibID")
  ) %>% 
  select(!LibID: x_axis) %>% 
  column_to_rownames("id")
  
```

```{r}
post_id_cluster_order <- metadata_tax_post %>% 
  mutate(
    id = factor(
      id,
      levels = str_subset(hclust_ward$labels[ward_order], "do", negate = T)
    )
  )


post_id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
         value = if_else(value < 10, round(value, 1), floor(value))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001))


ggsave("../figures/heatmap_post_hellinger.png", device = "png", width = 9, height = 10)

```



```{r}
gg_heat_post <- post_id_cluster_order %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  mutate(clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
         sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
         value = if_else(value < 10, round(value, 1), floor(value))) %>% 
  ggplot(aes(x = id, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = -0.0001),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```



```{r}
ggarrange(gg_heat_inc, gg_heat_post, ncol = 2, widths = c(3,2))

ggsave("../figures/heatmap_pre_post_hellinger.png", device = "png", width = 21, height = 10)
```



## NP heatmap 
```{r}
labels <- c("Pre", paste0("Day ", c(5,10,16,28)), "Post", paste0("Follow-up ", c(1,3,6, 12), "M"), "Donor")

metadata_weigth %>% 
  filter(clade_name %in% top_genera_weighted$clade_name) %>% 
  group_by(id, x_axis, clade_name) %>% 
  summarise(hellinger = mean(hellinger),
            value = mean(value)) %>% 
  mutate(
     clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
     x_axis = factor(x_axis, levels = stages, labels = labels ),
     value = if_else(value < 10, round(value, 1), floor(value)),
     id = factor(id, levels = c( "pt001", "pt002", "pt006", "asym004"))
     ) %>%
  ggplot(aes(x = x_axis, y = fct_rev(clade_name),  fill = hellinger, label = value)) +
    geom_tile(color = "white") +
    scale_fill_gradient(low = "white", high = "red") +
    geom_text(color = "black") +
    facet_grid(. ~ id, scales = "free_x", space = "free") +
    labs(x = "", y= "") +
    plot_theme +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.35, hjust = 1),
          axis.text = element_text(size = 12))

ggsave("../figures/heatmap_NP_hellinger.png", device = "png", width = 15, height = 10)
```






