---
title: "heatmap"
author: "Sebastian Mølvang Dall"
date: '2022-08-10'
output: html_document
---

## Loading libraries

```{r}
library(tidyverse)
library(ggpubr)
library(mplibrary)
source("./src/functions/metadata_functions.R")
source("./src/functions/gg_functions.R")
```

## Loading Data

**Important!:The MicroPouchNP heatmap is in the bottom. Follow the headers**

```{r}
metadata <- load_metadata()

metaphlan <- load_metaphlan(taxonomic_level = "genus")

project_filter <- "MP"
```

```{r}
metadata_in_metaphlan <- metadata %>%
  filter(LibID %in% colnames(metaphlan))


metaphlan_long <- metaphlan %>%
  select(-NCBI_tax_id) %>%
  pivot_longer(-clade_name, names_to = "LibID") %>%
  arrange(LibID)
```

### Filtering for relevant patients and donors

Just as in alpha_beta_diversity.rmd, the donors are filtered such that only donors used for the available patients are used.

```{r}
stages <- stages_pre_post


metadata_relevant <- isolateProjectMetadata(metadata_in_metaphlan, project_filter = "MP") %>%
  mutate(
    sample_group = paste0(x_axis, " ", group),
    sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group)
  ) %>%
  filter(x_axis %in% stages) %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2 | x_axis == "Donor")

n_genera <- 25
```

#### Adding taxonomy to metadata table

```{r}
metadata_relevant_with_tax <- left_join(metadata_relevant, metaphlan_long)
```

## heatmap visualization

### Donor, Pre, heatmap

#### Top genera
This chunk finds the most abundant genera by calculating the weighted mean of each taxonomic clade in each group, `Pre FMT`, `Pre Placebo`, and `Donor`. Each group is weighted by the inverse number of samples in each group. Since there are the least amount of samples in the `Donor` group their relative abundance will have a higher weight in the mean. The purpose is to find engraftment from the donor microbiome to the patients therefore the `Donor` microbiome gets a higher weight. In the end the top genera will be arranged according to their prevalence in the `Donor` group. 
```{r}
metadata_tax_hellinger <- hellingerTransformSamples(metadata_relevant_with_tax)
top_genera_weighted <- calculateTopGenera(metadata_tax_hellinger, n_genera)
top_genera_weighted_arranged_by_donor <- arrangeTopGeneraInDonor(metadata_tax_hellinger, top_genera_weighted)
```

#### hclust

For better visualization purposes the samples pre treatment are clustered based on their microbiome. Here a distance matrix using Bray-Curtis is calculated on Hellinger transformed relative abundances and clustered using the "ward D.2" algorithm, which tries to minimize within cluster variance using sum of squared(D2) errors.

```{r}
metadata_tax_clustered <- clusterPreAndDonorSamples(metadata_tax_hellinger, t_metaphlan)
```

## MP heatmaps
### Donor / Pre heatmap

Pre and donor heatmap clustered with hclust ward.D2 method. donor batches have been grouped by average relative abundance.

```{r}

gg_heat_inc <- metadata_tax_clustered %>%
  filter(clade_name %in% top_genera_weighted$clade_name) %>%
  mutate(
    clade_name = factor(clade_name, levels = top_genera_weighted_arranged_by_donor$clade_name),
    x_axis = factor(x_axis, levels = c("Pre", "Donor")),
    sample_group = factor(sample_group, levels = c("Pre FMT", "Pre placebo", "Donor")),
    relative_abundance = if_else(relative_abundance < 10, round(relative_abundance, 1), floor(relative_abundance))
  ) %>%
  ggplot(aes(x = id, y = fct_rev(clade_name), fill = hellinger, label = relative_abundance)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
  labs(x = "", y = "") +
  plot_theme +
  theme(
    axis.text.x = element_text(angle = 90, vjust = -0.0001),
    legend.position = "none"
  )
gg_heat_inc

# ggsave("../figures/heatmap_pre_donor_weight-hellinger.png", device = "png", width = 12, height = 10)
```



### Post heatmap


```{r}

metadata_tax_post <- metadata_tax_hellinger %>%
  filter(x_axis == "Post") %>%
  arrange(LibID)



t_metaphlan_post_id <- metadata_tax_post %>%
  select(id, LibID, group, sample_group, x_axis) %>%
  distinct(LibID, .keep_all = T) %>%
  left_join(
    rownames_to_column(t_metaphlan, "LibID")
  ) %>%
  select(!LibID:x_axis) %>%
  column_to_rownames("id")
```

```{r}
post_id_cluster_order <- metadata_tax_post %>%
  mutate(
    id = factor(
      id,
      levels = str_subset(hclust_ward$labels[ward_order], "do", negate = T)
    )
  )


post_id_cluster_order %>%
  filter(clade_name %in% top_genera_weighted$clade_name) %>%
  mutate(
    clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
    sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
    value = if_else(value < 10, round(value, 1), floor(value))
  ) %>%
  ggplot(aes(x = id, y = fct_rev(clade_name), fill = hellinger, label = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
  labs(x = "", y = "") +
  plot_theme +
  theme(axis.text.x = element_text(angle = 90, vjust = -0.0001))


ggsave("../figures/heatmap_post_weight-hellinger.png", device = "png", width = 9, height = 10)
```


This aligns Pre and post heatmap for visualisation.
```{r}
gg_heat_post <- post_id_cluster_order %>%
  filter(clade_name %in% top_genera_weighted$clade_name) %>%
  mutate(
    clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
    sample_group = factor(sample_group, levels = c("Post FMT", "Post placebo")),
    value = if_else(value < 10, round(value, 1), floor(value))
  ) %>%
  ggplot(aes(x = id, y = fct_rev(clade_name), fill = hellinger, label = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
  labs(x = "", y = "") +
  plot_theme +
  theme(
    axis.text.x = element_text(angle = 90, vjust = -0.0001),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
#
# ggarrange(gg_heat_inc, gg_heat_post, ncol = 2, widths = c(3,2))
#
# ggsave("../figures/heatmap_pre_post_hellinger.png", device = "png", width = 21, height = 10)
```

### Group by treatment

```{r}
id_cluster_order <- t_metaphlan_long_id_donor_avg %>%
  right_join(metadata_weigth) %>%
  mutate(
    id = factor(
      id,
      levels = hclust_ward$labels[ward_order]
    )
  ) %>%
  group_by(id, clade_name, sample_group, x_axis, group) %>%
  summarise(
    value = mean(value),
    hellinger = mean(hellinger)
  ) %>%
  filter(clade_name %in% top_genera_weighted$clade_name) %>%
  mutate(
    clade_name = factor(clade_name, levels = top_genera_weighted_donor$clade_name),
    x_axis = factor(x_axis, levels = c("Pre", "Post", "Donor")),
    sample_group = factor(sample_group, levels = c("Pre FMT", "Post FMT", "Pre placebo", "Post placebo", "Donor")),
    relative_abundance = if_else(value < 10, round(value, 1), floor(value))
  )



gg_heat_fmt <- id_cluster_order %>%
  filter(group == "FMT") %>%
  ggplot(aes(x = id, y = fct_rev(clade_name), fill = hellinger, label = relative_abundance)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
  labs(x = "", y = "") +
  plot_theme +
  theme(
    axis.text.x = element_text(angle = 90, vjust = -0.0001),
    legend.position = "none"
  )
gg_heat_fmt

ggsave("../figures/heatmap_fmt_donor.png", device = "png", width = 12, height = 10)
```


```{r}
gg_heat_placebo <- id_cluster_order %>%
  filter(group == "placebo") %>%
  ggplot(aes(x = id, y = fct_rev(clade_name), fill = hellinger, label = relative_abundance)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  geom_text(color = "black") +
  facet_grid(. ~ sample_group, scales = "free_x", space = "free") +
  labs(x = "", y = "") +
  plot_theme +
  theme(
    axis.text.x = element_text(angle = 90, vjust = -0.0001),
    legend.position = "none"
  )

gg_heat_placebo
ggsave("../figures/heatmap_placebo.png", device = "png", width = 9, height = 10)
```




