---
title: "Ordination plot"
author: "Sebastian Mølvang Dall"
date: '2022-08-24'
output: html_document
editor_options: 
  chunk_output_type: console
---


## libraries and functions

```{r}
library(tidyverse)
library(ampvis2)
library(RColorBrewer)
library(mplibrary)
library(RVAideMemoire)
library(gginnards)
source("./functions/metadata_functions.R")
source("./functions/bioinformatics.R")
```


## Load data

```{r}
metadata <- load_metadata()

metaphlan <- read_delim(
  "../data/metaphlan4/MetaPhlAn_4.0.3_Combined_NonHuman_Subsampled_2500000_profile.txt",
  delim = "\t",
  show_col_types = FALSE,
  skip = 1
) %>%
  filter_taxonomy("species", remove_tax_names = FALSE) %>%
  remove_NonHuman_from_colnames()

project_filter <- "MP"
```



### Prepare metadata

```{r}
metadata_MP <- metadata %>%
  isolateProjectMetadata(project_filter = "MP") %>%
  filter(sample_barcode %in% colnames(metaphlan)) %>%
  relocate(sample_barcode) %>%
  mutate(
    sample_group = paste0(x_axis, " ", group),
    sample_group = if_else(str_detect(sample_group, "Donor FMT"), str_remove(sample_group, " FMT"), sample_group)
  )
```


### Prepare tax data

```{r}
metaphlan_otu <- metaphlan %>%
  mutate(OTU = paste0("OTU", row_number())) %>%
  relocate(OTU)


otutable <- metaphlan_otu %>%
  select(-clade_name)


taxtable <- metaphlan_otu %>%
  select(OTU, clade_name) %>%
  mutate(
    clade_name = if_else(
      clade_name == "UNCLASSIFIED",
      "k__UNCLASSIFIED|p__UNCLASSIFIED|c__UNCLASSIFIED|o__UNCLASSIFIED|f__UNCLASSIFIED|g__UNCLASSIFIED|s__UNCLASSIFIED",
      clade_name
    )
  ) %>%
  separate(clade_name, sep = "\\|", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%
  relocate(OTU, .after = "Species")
```


## Load ampvis


```{r}
d <- amp_load(
  otutable = otutable,
  metadata = metadata_MP,
  taxonomy = taxtable
)
```


### Filter ampvis for pre post

```{r}
stages <- stages_pre_post

metadata_pre_post <- metadata_MP %>%
  filter(x_axis %in% stages, x_axis != "Donor") %>%
  group_by(id) %>%
  filter(n_distinct(x_axis) == 2) %>%
  ungroup()
```

```{r}
d_pre_post <- d %>%
  amp_subset_samples(sample_barcode %in% metadata_pre_post$sample_barcode)
```

# MP
### Ordination

```{r}
# shuffle barcodes
pca <- d %>%
  amp_subset_samples(sample_barcode %in% metadata_pre_post$sample_barcode | x_axis == "Donor") %>%
  amp_subset_taxa("k__UNCLASSIFIED", remove = TRUE) %>%
  amp_ordinate(
    type = "PCA",
    transform = "hellinger",
    sample_color_by = "sample_group",
    # sample_shape_by = "x_axis",
    sample_point_size = 4,
    species_plot = T,
    species_nlabels = 5,
    species_label_taxonomy = "Genus",
    filter_species = 0,
    # sample_plotly = "id",
    sample_colorframe = F,
    detailed_output = T
  )

meta <- pca$dsites %>%
  mutate(PC1 = -PC1)
meta_labels <- meta %>%
  filter(id == "pt014" & x_axis == "Post" | donor_batch == "do016_4")

species <- pca$dspecies %>%
  mutate(PC1 = -PC1)
species_labels <- species %>%
  filter(row_number() %in% 1:5) %>%
  rename(label = Genus)

model <- pca$model


format_theme <- theme(
  legend.title = element_blank(),
  panel.grid.major = element_line(color = "gray90", linewidth = 0.5),
  panel.grid.minor = element_line(color = "gray90", linewidth = 0.5),
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 6, face = "bold"),
  legend.text = element_text(size = 6), # Decrease legend text size
  legend.key.size = unit(0.5, "lines"), # Decrease legend key size
)

pca_plot <- ggplot(species, aes(PC1, PC2)) +
  geom_point(color = "grey70", size = 1) +
  geom_line(data = meta, aes(group = id), color = "grey70") +
  geom_point(data = meta, aes(color = sample_group), size = 2) +
  scale_color_brewer(palette = "Paired", labels = c("Donor", "After FMT", "After Placebo", "Before FMT", "Before Placebo")) +
  ggrepel::geom_text_repel(data = species_labels, aes(label = label), fontface = "bold.italic", size = 2) +
  ggrepel::geom_text_repel(data = meta_labels, aes(label = id), color = "red", size = 2) +
  theme_classic() +
  labs(
    x = paste0("PC1 [", round(model$CA$eig[1] / model$tot.chi * 100, 1), "%]"),
    y = paste0("PC2 [", round(model$CA$eig[2] / model$tot.chi * 100, 1), "%]")
  ) +
  format_theme



# ggsave("../figures/metaphlan4/ordination/pca_hellinger_pre_post_donor.png", device = "png", width = 7, height = 5)
```

```{r}
paired_2_5 <- brewer.pal(n = 5, "Paired")[2:5]

d_pre_post$metadata <- mutate(d_pre_post$metadata, sample_group = factor(
  sample_group,
  levels = c("Pre FMT", "Post FMT", "Pre placebo", "Post placebo"),
  labels = c("Before FMT", "After FMT", "Before Placebo", "After Placebo")
))

rda_ord_pt <- d_pre_post %>%
  amp_subset_taxa("k__UNCLASSIFIED", remove = TRUE) %>%
  amp_ordinate(
    type = "rda",
    constrain = "sample_group",
    transform = "hellinger",
    sample_color_by = "sample_group",
    sample_colorframe = T,
    sample_colorframe_label = "sample_group",
    species_plot = F,
    envfit_show = T,
    detailed_output = T,
    sample_colorframe_label_size = 2
  )


rda_plot <- rda_ord_pt$plot +
  labs(fill = "", color = "") +
  scale_color_manual(values = paired_2_5) +
  scale_fill_manual(values = paired_2_5) +
  theme_classic() +
  format_theme
rda_plot


ggsave("../figures/metaphlan4/ordination/amp_rda_hellinger_no_donor.png", device = "png", width = 7, height = 5)
```

```{r}
ggpubr::ggarrange(pca_plot, rda_plot, ncol = 2, font.label = list(size = 8, color = "black", face = "bold", family = NULL), labels = c("A", "B"))
ggsave("../figures/metaphlan4/ordination/amp_rda_pca_hellinger_no_donor.png", device = "png", width = 169, height = 63, units = "mm", dpi = 450)
```



```{r}
metadata_all_stages <- metadata_MP %>%
  filter(project %in% c("MP", "donor_batch"), x_axis %in% stages_all, group != "placebo") %>%
  mutate(x_axis = factor(x_axis, levels = stages_all)) %>%
  arrange(id, fecal_donation_number)

d_MP_all <- amp_load(
  otutable = otutable,
  metadata = metadata_all_stages,
  taxonomy = taxtable
)



d_MP_all %>%
  amp_ordinate(
    type = "pca",
    transform = "hellinger",
    sample_color_by = "id",
    sample_shape_by = "project",
    # sample_plotly = "donor_batch",
    species_plot = F,
    species_nlabels = 5,
    species_label_taxonomy = "Genus",
    filter_species = 0,
    sample_colorframe = F
  ) +
  geom_path(aes(group = id), color = "grey70")
```



Next chunck calculates the difference between centroids for groups in the PCA.
`envfit()` calculates the regression of supplementary variables on the ordination axes of unconstrained ordination. Furthermore `envfit()` can test the significance of this test by permutation (that is shuffling of the supplementary variable).

For factors significance of the fit is calculated by testing if the variation within centroids are significantly different from total spread of the data. That is $$r^{2} = 1 - \frac{ss_w}{ss_t} $$, where $$ss_{w}$$ and $$ss_t$$ are within-group and total sums of squares, respectively. To test if this fit is significant `envfit()` and `pairwise.factorfit()` uses permutations of the data to test if this explanation of variance is significant.

*It is crucial* that test for significance using `envfit()` is done on unconstrained ordination. If the test is done on constrained data the ordination axis already contain the information of the variable and as a consequence a randomly assigned variable can be highly significant (because the axis already contain the information).



```{r}
t_metaphlan <- metaphlan %>%
  transposeMetaphlan() %>%
  rownames_to_column("sample_barcode") %>%
  filter(sample_barcode %in% metadata_pre_post$sample_barcode) %>%
  arrange(sample_barcode) %>%
  column_to_rownames("sample_barcode") %>%
  abundance_filter(abundance_threshold = 10^-34) %>%
  vegan::decostand("hellinger") %>%
  rownames_to_column("sample_barcode")



MetadataMetaphlan <- metadata_pre_post %>%
  dplyr::select(sample_barcode) %>%
  ungroup() %>%
  left_join(t_metaphlan) %>%
  column_to_rownames("sample_barcode") %>%
  as.data.frame()

constrain <- metadata_pre_post %>%
  arrange(sample_barcode) %>%
  select(sample_group)


# RDA <- rda(MetadataMetaphlan ~ sample_group, data = constrain)
# envfit(RDA, env = constrain)

pca <- rda(MetadataMetaphlan)
envfit(pca ~ constrain$sample_group)

pairwise.factorfit(pca, constrain$sample_group)
```


## MP + Asym

```{r}
metadata <- metadata %>%
  filter(sample_barcode %in% colnames(metaphlan)) %>%
  relocate(sample_barcode)


metaphlan_otu <- metaphlan %>%
  # select(-NCBI_tax_id) %>%
  select(clade_name, metadata$sample_barcode) %>%
  mutate(OTU = paste0("OTU", row_number())) %>%
  relocate(OTU)

otutable <- metaphlan_otu %>%
  # select(-clade_name)

  taxtable() <- metaphlan_otu %>%
  select(OTU, clade_name) %>%
  separate(clade_name, sep = "\\|", into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")) %>%
  relocate(OTU, .after = "Species")
```


```{r}
metadata_sub <- metadata %>%
  filter(LibID %in% metadata_MP$LibID | id == "asym004") %>%
  filter(stage %in% c(NA, "inclusion", "followup_30d"), project != "donor_screening", group %in% c(NA, "FMT")) %>%
  mutate(
    project = if_else(project == "donor_batch",
      if_else(str_detect(id, "asym"), "Asymp Pouch", "Healthy Donor"), project
    ),
    stage = if_else(!is.na(stage), stage, "Donor"),
    project_stage = if_else(project == "MP", paste0(project, " ", stage), project)
  ) %>%
  group_by(id) %>%
  filter(n() > 1)
```


```{r}
d <- amp_load(
  otutable = otutable,
  metadata = metadata_sub,
  taxonomy = taxtable
)
```


```{r}
pca <- d %>%
  amp_ordinate(
    type = "pca",
    transform = "hellinger",
    sample_color_by = "project_stage",
    sample_shape_by = "stage",
    species_plot = T,
    species_nlabels = 5,
    species_label_taxonomy = "Genus",
    filter_species = 0,
    sample_colorframe = F,
    sample_point_size = 3
  ) +
  geom_line(aes(group = id), color = "grey70") +
  theme(legend.title = element_blank()) +
  scale_color_brewer(palette = "Dark2") +
  annotate(geom = "text", x = 0, y = 0.53, label = "Asymptomatic Pouch", color = "#1a401a") +
  annotate(geom = "text", x = -0.3, y = -0.35, label = "Healthy Donors", color = "#955612")
pca

ggsave("../figures/amp_pca_hellinger_fmt_donor.png", device = "png", width = 7, height = 5)
```


